05_DESeq_manipulation

### 2019-05-19 analysis/05_DESeq2_manipulate.Rmd
cleaning up, prep for separarate functions

```{r setup, message=F, comment=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(DESeq2)
library(cowplot)
library(RColorBrewer)
library(pheatmap)
library(kableExtra)
library(viridis)

@@

knitr::opts_chunk$set(fig.path = '../figures/manipulation/')
```

# Manipulation data

@@

# Write for loop to do this one for every tissue and for every treatment

```{r althings, eval = T}
## subset for test 
#m.colData <- m.colData %>%
#  filter(sextissue == "female_pituitary") %>%
#  droplevels()

@@

  dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ treatment )
  dds <- dds[ rowSums(counts(dds)) > 2, ] ## pre-filter genes 
  dds <- DESeq(dds) # Differential expression analysis
  vsd <- vst(dds, blind=FALSE) # variance stabilized

@@

# comapre all contrasts, save to datafrmes
dat=data.frame()
for (i in a){
  for (j in b){
    if (i != j) {
      k <- paste(i,j, sep = "") #assigns usique rownames
      dat[k,1]<-i               
      dat[k,2]<-j
      dat[k,3]<- numDEGs(i,j) #caluculates number of DEGs
    }
  }
}

head(dat)

# widen data to create table of degs
rownames(dat) <- NULL #remove row names
data_wide <- spread(dat, V2, V3)
print(data_wide)

dat$V1 <- factor(dat$V1, levels = 
                              c("m.inc.d3",  "m.inc.d8",
                                "m.inc.d9", "m.inc.d17",
                                "prolong", "extend", "m.n2"))
dat$V2 <- factor(dat$V2, levels = 
                              c("m.inc.d3",  "m.inc.d8",
                                "m.inc.d9", "m.inc.d17",
                                "prolong", "extend", "m.n2"))

allcontrasts <- dat %>%
  ggplot( aes(V1, V2)) +
    geom_tile(aes(fill = V3)) +
    scale_fill_viridis(na.value="#FFFFFF00", 
                     limits = c(0, 6000),
                     breaks = c(0, 1000, 2000, 3000, 4000, 5000, 6000)) + 
    xlab(" ") + ylab("Timepoint") +
    labs(fill = "# of DEGs",
         subtitle = eachgroup)
plot(allcontrasts)

# create the dataframe using my function pcadataframe
pcadata <- pcadataframe(vsd, intgroup=c("treatment"), returnData=TRUE)
percentVar <- round(100 * attr(pcadata, "percentVar"))
percentVar

pca12 <- ggplot(pcadata, aes(PC1, PC2,color = treatment)) + 
  geom_point(size = 2, alpha = 1) +
  stat_ellipse(type = "t") +
  xlab(paste0("PC1: ", percentVar[1],"% variance")) +
  ylab(paste0("PC2: ", percentVar[2],"% variance")) +
  theme_cowplot(font_size = 8, line_size = 0.25) +
  labs(subtitle = eachgroup)
print(pca12)

@@

print(summary(aov(PC2 ~ treatment, data=pcadata))) 
print(TukeyHSD(aov(PC2 ~ treatment, data=pcadata), which = "treatment"))

pca34 <- ggplot(pcadata, aes(PC3, PC4,color = treatment)) + 
  geom_point(size = 2, alpha = 1) +
  stat_ellipse(type = "t") +
  xlab(paste0("PC3: ", percentVar[3],"% variance")) +
  ylab(paste0("PC4: ", percentVar[4],"% variance")) +
  theme_cowplot(font_size = 8, line_size = 0.25) +
  labs(subtitle = eachgroup)
#print(pca34)

# see http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#heatmap-of-the-count-matrix
sampleDists <- dist(t(assay(vsd)))

sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- NULL
colnames(sampleDistMatrix) <- colData$treatment
colors <- colorRampPalette(brewer.pal(9, "Blues"))(255)

#pheatmap(sampleDistMatrix,
#         clustering_distance_rows=sampleDists,
#         clustering_distance_cols=sampleDists,
#         col=colors,
 #        fontsize = 6, 
 #        main = eachgroup)

# Multi-factor designs
# revals is my function easily calculate multiple multi factor designs
# it is built around the documentaion from DESeq which looks more simply like
# res <- results(dds,contrast=c("columnname", "factor1", "factor2"))
# then I extract the p-values from within the results for showing number of DEGes
resvals <- function(contrastvector, mypval){
  res <- results(dds, contrast = c(contrastvector[1],contrastvector[2],contrastvector[3]), independentFiltering = T)
  sumpvalue <- sum(res$pvalue < mypval, na.rm = TRUE)
  #print(sumpvalue)
  sumpadj <- sum(res$padj < mypval, na.rm = TRUE)
  print(sumpadj)
  vals <- cbind(res$pvalue, res$padj)
  pvalcolname <- as.character(paste("pval",contrastvector[1],contrastvector[2],contrastvector[3], sep=""))
  padjcolname <- as.character(paste("padj",contrastvector[1],contrastvector[2],contrastvector[3], sep=""))
  colnames(vals) <- c(pvalcolname, padjcolname)
  return(vals)
}
## DEG by contrasts at 0.1 pvalue
contrast1 <- resvals(contrastvector = c('treatment', 'm.inc.d3', 'm.inc.d9'), mypval = 0.01) #12
contrast2 <- resvals(contrastvector = c('treatment', 'm.inc.d3', 'm.inc.d17'), mypval = 0.01) #73
contrast3 <- resvals(contrastvector = c('treatment', 'm.inc.d3', 'm.n2'), mypval = 0.01) #284
contrast4 <- resvals(contrastvector = c('treatment', 'm.inc.d3', 'm.inc.d8'), mypval = 0.01) #15
contrast5 <- resvals(contrastvector = c('treatment', 'm.inc.d3', 'prolong'), mypval = 0.01) #555
contrast6 <- resvals(contrastvector = c('treatment', 'm.inc.d3', 'extend'), mypval = 0.01) #1109
contrast7 <- resvals(contrastvector = c('treatment', 'm.inc.d9', 'm.inc.d17'), mypval = 0.01) #280
contrast8 <- resvals(contrastvector = c('treatment', 'm.inc.d9', 'm.n2'), mypval = 0.01) #440
contrast9 <- resvals(contrastvector = c('treatment', 'm.inc.d9', 'm.inc.d8'), mypval = 0.01) #484
contrast10 <- resvals(contrastvector = c('treatment', 'm.inc.d9', 'prolong'), mypval = 0.01) #529
contrast11 <- resvals(contrastvector = c('treatment', 'm.inc.d9', 'extend'), mypval = 0.01) #1177

contrast12 <- resvals(contrastvector = c('treatment', 'm.inc.d17', 'm.n2'), mypval = 0.01) #5
contrast13 <- resvals(contrastvector = c('treatment', 'm.inc.d17', 'm.inc.d8'), mypval = 0.01) #553
contrast14 <- resvals(contrastvector = c('treatment', 'm.inc.d17', 'prolong'), mypval = 0.01) #339
contrast15 <- resvals(contrastvector = c('treatment', 'm.inc.d17', 'extend'), mypval = 0.01) #782
contrast16 <- resvals(contrastvector = c('treatment', 'm.n2', 'm.inc.d8'), mypval = 0.01) #1123
contrast17 <- resvals(contrastvector = c('treatment', 'm.n2', 'prolong'), mypval = 0.01) #265
contrast18 <- resvals(contrastvector = c('treatment', 'm.n2', 'extend'), mypval = 0.01) #552
contrast19 <- resvals(contrastvector = c('treatment', 'm.inc.d8', 'prolong'), mypval = 0.01) #828
contrast20 <- resvals(contrastvector = c('treatment', 'm.inc.d8', 'extend'), mypval = 0.01) #1110
contrast21 <- resvals(contrastvector = c('treatment', 'prolong', 'extend'), mypval = 0.01) #64

DEGs <- assay(vsd)
DEGs <- cbind(DEGs, contrast1, contrast2, contrast3, contrast4, contrast5,
              contrast6, contrast7, contrast8, contrast9, contrast10,
              contrast11, contrast12, contrast13, contrast14, contrast15,
              contrast16, contrast17, contrast18, contrast19, contrast20,contrast21)
DEGs <- as.data.frame(DEGs) # convert matrix to dataframe
DEGs$rownames <- rownames(DEGs)  # add the rownames to the dataframe

#DEGs[is.na(DEGs)] <-  1

DEGs$padjmin <- with(DEGs, pmin(pvaltreatmentm.inc.d3m.inc.d9, 
  padjtreatmentm.inc.d3m.inc.d9,               
  pvaltreatmentm.inc.d3m.inc.d17, padjtreatmentm.inc.d3m.inc.d17,              
  pvaltreatmentm.inc.d3m.n2, padjtreatmentm.inc.d3m.n2,                  
  pvaltreatmentm.inc.d3m.inc.d8, padjtreatmentm.inc.d3m.inc.d8,               
  pvaltreatmentm.inc.d3prolong, padjtreatmentm.inc.d3prolong,               
  pvaltreatmentm.inc.d3extend, padjtreatmentm.inc.d3extend,                
  pvaltreatmentm.inc.d9m.inc.d17, padjtreatmentm.inc.d9m.inc.d17,             
  pvaltreatmentm.inc.d9m.n2, padjtreatmentm.inc.d9m.n2,                  
  pvaltreatmentm.inc.d9m.inc.d8, padjtreatmentm.inc.d9m.inc.d8,              
  pvaltreatmentm.inc.d9prolong, padjtreatmentm.inc.d9prolong,               
  pvaltreatmentm.inc.d9extend, padjtreatmentm.inc.d9extend,                
  pvaltreatmentm.inc.d17m.n2, padjtreatmentm.inc.d17m.n2,                 
  pvaltreatmentm.inc.d17m.inc.d8, padjtreatmentm.inc.d17m.inc.d8,             
  pvaltreatmentm.inc.d17prolong, padjtreatmentm.inc.d17prolong,
  pvaltreatmentm.inc.d17extend,padjtreatmentm.inc.d17extend, 
  pvaltreatmentm.n2m.inc.d8, padjtreatmentm.n2m.inc.d8,                  
  pvaltreatmentm.n2prolong, padjtreatmentm.n2prolong,                   
  pvaltreatmentm.n2extend, padjtreatmentm.n2extend,                    
  pvaltreatmentm.inc.d8prolong, padjtreatmentm.inc.d8prolong,               
  pvaltreatmentm.inc.d8extend, padjtreatmentm.inc.d8extend,                
  pvaltreatmentprolongextend, padjtreatmentprolongextend))

sigDEGs <- DEGs %>% arrange(padjmin)
sigDEGs <- head(sigDEGs,500)
sigDEGs <- as.data.frame(sigDEGs)
rownames(sigDEGs) <- sigDEGs$rownames
drop.cols <-colnames(sigDEGs[,grep("padj|pval|pmin|rownames", colnames(sigDEGs))])
sigDEGs <- sigDEGs %>% dplyr::select(-one_of(drop.cols))
sigDEGs <- as.matrix(sigDEGs)
sigDEGs <- sigDEGs - rowMeans(sigDEGs)

@@

sigDEGs <- as.matrix(sigDEGs) 
pheatmap(sigDEGs, show_rownames = F, show_colnames = F,
         color = viridis(30),
         breaks=myBreaks,
         annotation_col=anndf,
         main = eachgroup)

pheatmap(sigDEGs, kmeans_k = 5,
         show_rownames = F, show_colnames = F,
         color = viridis(30),
         breaks=myBreaks,
         annotation_col=anndf,
         main = eachgroup)
}
```

```{r allthings-outcome, eval = T}

  dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ outcome )
  dds <- dds[ rowSums(counts(dds)) > 2, ] ## pre-filter genes 
  dds <- DESeq(dds) # Differential expression analysis
  vsd <- vst(dds, blind=FALSE) # variance stabilized 
head(vsd)

numDEGs <- function(group1, group2){
  res <- results(dds, contrast = c("outcome", group1, group2), independentFiltering = T)
  sumpadj <- sum(res$padj < 0.1, na.rm = TRUE)
  return(sumpadj)}

#create list of groups
a <- levels(colData$outcome)
b <- levels(colData$outcome)

# comapre all contrasts, save to datafrmes
dat=data.frame()
for (i in a){
  for (j in b){
    if (i != j) {
      k <- paste(i,j, sep = "") #assigns usique rownames
      dat[k,1]<-i               
      dat[k,2]<-j
      dat[k,3]<- numDEGs(i,j) #caluculates number of DEGs
    }
  }
}

head(dat)

# widen data to create table of degs
rownames(dat) <- NULL #remove row names
data_wide <- spread(dat, V2, V3)
print(data_wide)

dat$V1 <- factor(dat$V1, levels = 
                              c("end inc",  "end hatch",
                                "prolong inc", "delay hatch"))
dat$V2 <- factor(dat$V2, levels = 
                              c("end inc",  "end hatch",
                                "prolong inc", "delay hatch"))

allcontrasts <- dat %>%
  ggplot( aes(V1, V2)) +
    geom_tile(aes(fill = V3)) +
    scale_fill_viridis(na.value="#FFFFFF00", 
                     limits = c(0, 6000),
                     breaks = c(0, 1000, 2000, 3000, 4000, 5000, 6000)) + 
    xlab(" ") + ylab("Timepoint") +
    labs(fill = "# of DEGs",
         subtitle = eachgroup)
plot(allcontrasts)

# create the dataframe using my function pcadataframe
pcadata <- pcadataframe(vsd, intgroup=c("outcome"), returnData=TRUE)
percentVar <- round(100 * attr(pcadata, "percentVar"))
percentVar

pca12 <- ggplot(pcadata, aes(PC1, PC2,color = outcome)) + 
  geom_point(size = 2, alpha = 1) +
  stat_ellipse(type = "t") +
  xlab(paste0("PC1: ", percentVar[1],"% variance")) +
  ylab(paste0("PC2: ", percentVar[2],"% variance")) +
  theme_cowplot(font_size = 8, line_size = 0.25) +
  labs(subtitle = eachgroup)
print(pca12)

print(summary(aov(PC1 ~ outcome, data=pcadata)))
print(TukeyHSD(aov(PC1 ~ outcome, data=pcadata), which = "outcome"))

print(summary(aov(PC2 ~ outcome, data=pcadata))) 
print(TukeyHSD(aov(PC2 ~ outcome, data=pcadata), which = "outcome"))

pca34 <- ggplot(pcadata, aes(PC3, PC4,color = outcome)) + 
  geom_point(size = 2, alpha = 1) +
  stat_ellipse(type = "t") +
  xlab(paste0("PC3: ", percentVar[3],"% variance")) +
  ylab(paste0("PC4: ", percentVar[4],"% variance")) +
  theme_cowplot(font_size = 8, line_size = 0.25) +
  labs(subtitle = eachgroup)
#print(pca34)

>>>>

```{r setup, message=F, comment=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(DESeq2)
library(cowplot)
library(RColorBrewer)
library(pheatmap)
library(kableExtra)
library(viridis)
library(edgeR)

@@

knitr::opts_chunk$set(fig.path = '../figures/manipulation/', cache = TRUE)
```

## Manipulation data

@@

## Make a PCA plot, heat map, and box plots for each tissue for each sex

```{r althings, eval = T}
## subset for test 
#m.colData <- m.colData %>%
#  filter(sextissue == "male_hypothalamus") %>%
# droplevels()

@@

  dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ treatment )

  print(dds)
  nrow(colData) > 50

  dds <- dds[rowSums(counts(dds) > 1) >= 10]  # filter more than sample with less 0 counts
  print(dds)

  dds <- DESeq(dds) # Differential expression analysis

  vsd <- vst(dds, blind=FALSE) # variance stabilized

@@

# comapre all contrasts, save to datafrmes
totalDEGS=data.frame()
for (i in a){
  for (j in b){
    if (i != j) {
      k <- paste(i,j, sep = "") #assigns usique rownames
      print(k)
      totalDEGS[k,1]<-i               
      totalDEGS[k,2]<-j
      totalDEGS[k,3]<- numDEGs(i,j) #caluculates number of DEGs
    }
  }
}

head(totalDEGS)

# widen data to create table of degs
rownames(totalDEGS) <- NULL #remove row names
totalDEGS_wide <- spread(totalDEGS, V2, V3)
print(totalDEGS_wide)

totalDEGS$V1 <- factor(totalDEGS$V1, levels = 
                              c("m.inc.d3",  "m.inc.d8",
                                "m.inc.d9", "m.inc.d17",
                                "prolong", "extend", "m.n2"))
totalDEGS$V2 <- factor(totalDEGS$V2, levels = 
                              c("m.inc.d3",  "m.inc.d8",
                                "m.inc.d9", "m.inc.d17",
                                "prolong", "extend", "m.n2"))

allcontrasts <- totalDEGS %>%
  ggplot( aes(V1, V2)) +
    geom_tile(aes(fill = V3)) +
    scale_fill_viridis(na.value="#FFFFFF00", 
                     limits = c(0, 3000),
                     breaks = c(0, 1000, 2000, 3000)) + 
    xlab(" ") + ylab("Timepoint") +
    labs(fill = "# of DEGs",
         subtitle = "eachgroup")
plot(allcontrasts)

# create the dataframe using my function pcadataframe
pcadata <- pcadataframe(vsd, intgroup=c("treatment"), returnData=TRUE)
percentVar <- round(100 * attr(pcadata, "percentVar"))
print(percentVar)

pca1 <- ggplot(pcadata, aes(treatment, PC1,color = treatment)) + 
  geom_boxplot() +
  ylab(paste0("PC1: ", percentVar[1],"% variance")) +
  xlab(NULL) +
  theme_cowplot(font_size = 8, line_size = 0.25) +
  labs(subtitle = "eachgroup") +
  theme(legend.position = "none")

pca2 <- ggplot(pcadata, aes(treatment, PC2,color = treatment)) + 
  geom_boxplot() +
  ylab(paste0("PC2: ", percentVar[2],"% variance")) +
  xlab(NULL) +
  theme_cowplot(font_size = 8, line_size = 0.25) +
  labs(subtitle = "eachgroup") +
  theme(legend.position = "none")

plot_grid(pca1, pca2)

@@

print(summary(aov(PC2 ~ treatment, data=pcadata))) 
print(summary(aov(PC3 ~ treatment, data=pcadata))) 
print(summary(aov(PC4 ~ treatment, data=pcadata)))

## heamap with minimum pvalue

# make dataframe counts
DEGs <- assay(vsd)
DEGs <- as.data.frame(DEGs)

# add pvalues

for (i in a){
  for (j in b){
    if (i != j) {
      k <- paste(i,j, sep = " vs ")
      print(k)
      results <- resvals2(i,j)
      DEGs <- cbind(DEGs,results)
    }
  }
}

DEGsmatrix <- DEGs
DEGsmatrix <- as.matrix(DEGs)
padjmin <- rowMins(DEGsmatrix, na.rm = T) 
padjmin <- as.data.frame(padjmin)

sigDEGs <- cbind(DEGs,padjmin)

sigDEGs <- sigDEGs %>% arrange(padjmin)
sigDEGs <- head(sigDEGs,500)
sigDEGs <- as.data.frame(sigDEGs)
rownames(sigDEGs) <- sigDEGs$rownames
drop.cols <-colnames(sigDEGs[,grep("padj|pval|pmin|rownames", colnames(sigDEGs))])
sigDEGs <- sigDEGs %>% dplyr::select(-one_of(drop.cols))
sigDEGs <- as.matrix(sigDEGs)
sigDEGs <- sigDEGs - rowMeans(sigDEGs)

@@

sigDEGs <- as.matrix(sigDEGs) 
pheatmap(sigDEGs, show_rownames = F, show_colnames = F,
         color = viridis(30),
         breaks=myBreaks,
         annotation_col=anndf,
         main = "eachgroup")

pheatmap(sigDEGs, kmeans_k = 5,
         show_rownames = F, show_colnames = F,
         color = viridis(30),
         breaks=myBreaks,
         annotation_col=anndf,
         main = "eachgroup")

## candidate genes
names(geneinfo)
names(geneinfo)[4] <- "rownames"
DEGs$rownames <- row.names(DEGs)

# make dataframe with geneids and names and counts

candidates <- full_join(geneinfo, DEGs)
drop.cols <-colnames(candidates[,grep("padj|pval|pmin", colnames(candidates))])
candidates <- candidates %>% dplyr::select(-one_of(drop.cols))
candidates <- candidates %>%
  filter(Name %in% c( #"JUN", "JUND", "EGR",  "AVP", "AVPR1A", "AVPR1B", "AVPR2", "OXT",  
                     "AR", "CYP19A1", "ESR1", "ESR2", "FSHR",
                     "GHRL", "GAL", "NPVF", "GNRH1", "LHCGR",
                     "PGR", "PRL", "PRLR", "VIP", "VIPR1")) 
row.names(candidates) <- candidates$Name
candidates <- candidates %>% select(-row.names, -rownames, -Name, -geneid)
candidates <- candidates %>% drop_na()

head(candidates)

candidates <- as.data.frame(t(candidates))
candidates$RNAseqID <- rownames(candidates)
head(candidates)

candidates <- candidates %>% gather(gene, value, -RNAseqID)  %>% # https://tidyr.tidyverse.org/reference/gather.html
  filter(RNAseqID != "gene")
candidates$value <- as.numeric(candidates$value)
candidates$V1  <- candidates$RNAseqID
head(candidates)

candidatecounts <- left_join(candidates, colData)
#head(candidatecounts)

candidatecounts$faketime <- as.numeric(candidatecounts$treatment)
#head(candidatecounts)

candidatecounts$gene <- as.factor(candidatecounts$gene)
levels(candidatecounts$gene)

p1 <- candidatecounts %>%
  filter(gene %in% c( "AR", "CYP19A1", "ESR1", "ESR2", "FSHR")) %>%
  ggplot(aes(x = treatment, y = value, fill = treatment)) +
  geom_boxplot() +
  facet_wrap(~gene, scales = "free", nrow = 1) +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
  labs(subtitle = "eachgroup")
p1

print(p1)

cat("\n")

p2 <- candidatecounts %>%
  filter(gene %in% c("GHRL", "GAL", "NPVF", "GNRH1", "LHCGR")) %>%
  ggplot(aes(x = treatment, y = value, fill = treatment)) +
  geom_boxplot() +
  facet_wrap(~gene, scales = "free", nrow = 1) +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
  labs(subtitle = "eachgroup")
p2

print(p2)

cat("\n")

p3 <- candidatecounts %>%
  filter(gene %in% c("PGR", "PRL", "PRLR", "VIP", "VIPR1")) %>%
  ggplot(aes(x = treatment, y = value, fill = treatment)) +
  geom_boxplot() +
  facet_wrap(~gene, scales = "free", nrow = 1) +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
  labs(subtitle = "eachgroup")
p3

print(p3)

cat("\n")

cat("Done with eachgroup")

cat("\n")
